---
title: "Three ways to state the same survival difference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Three ways to state the same survival difference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.2, fig.align = "center")
options(width = 88)
library(gistsurv)
```

A survival difference between two groups can be reported as a hazard ratio, as
a difference in survival probability at a fixed horizon, or as a difference in
restricted mean survival time. The three are not interchangeable. They can
differ in size, in statistical significance and, occasionally, in sign, and
which one you report changes what a reader takes away.

This vignette walks through a full analysis on `sim_gist`, a simulated cohort
that ships with the package, and shows where the three estimators agree, where
they part company, and what a competing-risk analysis says about why.

`sim_gist` is synthetic. Every parameter behind it is a literal in
`data-raw/make-sim-gist.R`; it is not a sample of any registry and no
epidemiological claim can be read off it. What is realistic about it is its
*shape* -- the variable names, the levels, the missingness, the awkward
cause-of-death coding -- which is what makes it a fair rehearsal for a real
extract.

## Preparing the outcome

Registry data arrives with a follow-up time, a vital status and a cause of
death, not with event indicators. `prep_surv()` builds them.

One step comes first. Registries commonly code cause of death with a value that
is carried by living subjects *and* by those who died of something else --
here, `"Alive or dead of other cause"`. That value cannot be declared to
`prep_surv()` as a cause of death, because the function refuses a declared
cause of death on a living subject, and it is right to: the same check is what
catches a genuine contradiction everywhere else. Resolve it first, against
vital status:

```r
raw$cause_of_death <- ifelse(
  raw$cause_of_death == "Alive or dead of other cause",
  ifelse(raw$vital_status == "Dead", "Other cause", NA),
  raw$cause_of_death)
```

`sim_gist` ships with those two lines already applied, so we can go straight
on. With the column unambiguous, every admissible value can be declared, and
an unexpected code stops the call instead of being folded into "death from
other causes".

```{r prep}
d <- prep_surv(
  sim_gist,
  time                = "time_mo",
  status              = "vital_status",
  cause               = "cause_of_death",
  cause_gist_value    = "Dead (attributable to this cancer dx)",
  cause_other_value   = "Other cause",
  cause_unknown_value = "Dead (missing/unknown COD)",
  cause_censor_value  = NA,
  cause_unknown_to    = 2L
)

attr(d, "os_cr_table")
```

Nothing is printed by the function itself. The cross-tabulation above is an
attribute, and it is the first thing to look at: the off-diagonal cells must be
empty, or the two outcome variables contradict each other. `prep_surv()`
enforces that before returning, so an empty off-diagonal here is a check that
ran, not a coincidence.

The counts include the figure a limitations paragraph has to quote -- how many
deaths carried no usable cause and were therefore assigned to a category
rather than observed in it:

```{r prep-counts}
unlist(attr(d, "prep_surv")$counts)
```

## Who is in the cohort

```{r table1}
t1 <- tab_baseline(
  d, by = "site",
  vars = c("age", "age_grp", "sex", "size_grp", "stage", "surgery")
)
head(t1[, c("label", "stat_label", "Overall", "Stomach", "Small intestine",
            "Colorectal", "smd_fmt")], 12)
```

The standardized mean difference is reported instead of a p-value. With more
than two groups it is the largest absolute SMD over all pairs, and `smd_pair`
names the pair, so one badly imbalanced pair is not diluted by the others.

## One contrast, three estimators

Take stomach against small intestine. First the Kaplan-Meier curves and the
median:

```{r km}
k <- fit_km(d, group = "site", times = c(12, 36, 60))
k[, c("stratum", "n", "events", "median_ci_txt", "median_fu_revkm", "max_fu")]
```

Median follow-up is the reverse Kaplan-Meier estimate, not the median of the
observed times: the latter is a summary of when people left the study, which
in a cohort with deaths is not the same question.

Then the hazard ratio, with the proportional-hazards test that has to
accompany it:

```{r cox}
cx <- fit_cox(d, covariates = c("site", "age_grp", "sex", "size_grp", "surgery"))
cx[cx$variable == "site", c("level", "n", "n_event", "hr_txt", "p_fmt")]

zph <- attr(cx, "cox_zph")
zph[, c("term", "chisq", "df", "p_fmt", "violates")]
```

A hazard ratio is a single number standing in for a whole curve. It is worth
that only if the ratio is roughly constant over time, which is what the test
above is for. `fit_cox()` will not return hazard ratios whose assumption was
never tested: if `cox.zph()` cannot produce a test -- as happens with
`transform = "log"` when there are events at time zero, which registry extracts
have -- the call stops rather than returning a missing test that reads like a
passed one.

Then the restricted mean survival time, which needs no proportionality
assumption because it is just an area under the curve:

```{r rmst}
r <- calc_rmst(d, group = "site", arms = c("Stomach", "Small intestine"),
               tau = c(12, 36, 60))
r[, c("tau", "rmst_arm0_txt", "rmst_arm1_txt", "rmst_diff_txt",
      "rmst_diff_p_fmt")]
```

It does need a horizon, and the horizon has to be one the data support. The
area up to `tau` is not defined past the largest follow-up time in either arm,
so `calc_rmst()` refuses rather than extrapolating:

```{r rmst-guard, error = TRUE}
calc_rmst(d, group = "site", arms = c("Stomach", "Colorectal"), tau = 400)
```

This guard is the reason to use the wrapper: `survRM2::rmst2()` on its own
will return a number there.

## The three side by side

`compare_estimands()` does all of the above for one contrast and lines the
answers up, with a verdict on whether they agree:

```{r compare}
ce <- compare_estimands(d, group = "site",
                        arms = c("Stomach", "Small intestine"),
                        tau = c(12, 36, 60))
ce[, c("tau", "hr_txt", "km_surv_diff_txt", "rmst_diff_txt", "verdict")]
```

The hazard ratio does not depend on `tau`, so it is repeated on every row; the
other two do, and are not. That is the first thing the table shows: two of the
three estimators answer a question that contains a horizon, and one does not.

Run it over several contrasts and the spread becomes visible.

```{r sweep}
contrasts <- list(
  c("site",    "Stomach",   "Small intestine"),
  c("site",    "Stomach",   "Colorectal"),
  c("age_grp", "<50",       "65-74"),
  c("age_grp", "<50",       "75+"),
  c("sex",     "Male",      "Female"),
  c("stage",   "Localized", "Regional")
)

sweep <- do.call(rbind, lapply(contrasts, function(p) {
  z <- compare_estimands(d, group = p[1], arms = p[2:3], tau = 60)
  data.frame(contrast = z$comparison, hr = z$hr_txt,
             km = z$km_surv_diff_txt, rmst = z$rmst_diff_txt,
             n_sig = z$n_signif, verdict = substr(z$verdict, 1, 46))
}))
sweep[, c("contrast", "n_sig", "verdict")]
```

Three of these six contrasts are discordant at 60 months, and they are
discordant in different ways. Male versus female is significant on the survival
difference and on the RMST difference but not on the hazard ratio. Localized
versus regional stage is the opposite: significant on the hazard ratio alone.
Reporting any one of the three and stopping would have produced a different
paper.

```{r sweep-plot, echo = FALSE}
op <- par(mar = c(4.2, 11, 2.4, 1.2))
h <- sapply(contrasts, function(p) {
  z <- compare_estimands(d, group = p[1], arms = p[2:3], tau = 60)
  c(z$rmst_diff, z$rmst_diff_lcl, z$rmst_diff_ucl)
})
lab <- sweep$contrast
y <- seq_along(lab)
plot(h[1, ], y, xlim = range(h, 0), ylim = c(0.5, length(y) + 0.5),
     pch = 19, yaxt = "n", ylab = "", bty = "n",
     xlab = "RMST difference at 60 months (months)",
     main = "Arm 1 minus arm 0")
segments(h[2, ], y, h[3, ], y)
abline(v = 0, lty = 3)
axis(2, at = y, labels = lab, las = 1, cex.axis = 0.8, tick = FALSE)
par(op)
```

## Why they disagree

A hazard ratio and an RMST difference can point different ways when the
competing risks differ between the groups: deaths from other causes remove
people from the risk set without being the event of interest, and the two
estimators absorb that differently.

`fit_competing()` splits the deaths by cause and estimates a cumulative
incidence for each.

```{r cif}
fc <- fit_competing(d, group = "age_grp", covariates = "age_grp",
                    times = c(12, 36, 60),
                    cause_labels = c("Disease", "Other causes"),
                    cause_specific = "skip")
full <- fc[fc$analysis_set == "full" & fc$time == 60, ]
full[, c("group", "cause_label", "n_group", "n_event_cause", "cif", "cif_lcl",
         "cif_ucl")]
```

The point of the table is the contrast between the two panels below. The
cumulative incidence of death from the disease moves around without much
pattern across the age groups. The cumulative incidence of death from other
causes climbs steadily and by 75 and over is the larger of the two. Most of
what an overall-survival analysis counts in that group is not the disease at
all.

```{r cif-plot, echo = FALSE, fig.height = 3.8}
op <- par(mfrow = c(1, 2), mar = c(4.2, 4.2, 2.6, 1))
for (cs in unique(full$cause_label)) {
  s <- full[full$cause_label == cs, ]
  b <- barplot(s$cif, names.arg = s$group, ylim = c(0, max(full$cif_ucl) * 1.1),
               xlab = "Age group", ylab = "Cumulative incidence at 60 months",
               main = cs, las = 1)
  arrows(b, s$cif_lcl, b, s$cif_ucl, angle = 90, code = 3, length = 0.05)
}
par(op)
```

The share of deaths that are not from the disease is the quantity an
overall-survival comparison silently averages over:

```{r share}
share <- do.call(rbind, lapply(split(full, full$group), function(s) {
  data.frame(age_grp = s$group[1],
             other_share_pct = round(
               100 * s$cif[s$cause_label == "Other causes"] / sum(s$cif), 1))
}))
share
```

Fine-Gray models put the same comparison on a regression footing. The
`covariates` argument is what asks for them; without it the call returns
cumulative incidences only, and `attr(fc, "finegray")` is `NULL`.

```{r finegray}
fg <- attr(fc, "finegray")
fg[fg$variable == "age_grp",
   c("cause_label", "level", "n", "n_event_cause", "shr_txt", "p_fmt")]
```

The two columns of subdistribution hazard ratios are the point. Across age the
disease-specific one barely moves; the other-cause one rises by an order of
magnitude. An overall-survival comparison adds those two together and reports
the sum as though it described the disease.

That is the mechanism behind the one discordant age contrast above. At 65-74
versus under 50 the hazard ratio and the survival difference at 60 months are
both significant while the RMST difference is not: the hazard ratio accumulates
the excess mortality over the whole of follow-up, where the other-cause
component has had time to dominate, whereas the restricted mean over a fixed 60
months weights the early part of the curve, where it has not.

## What the functions refuse to do

Every function in the package returns a data frame and prints nothing;
diagnostics come back as attributes. Several also stop rather than return a
number that would be read as an estimate:

- `prep_surv()` stops on a cause of death it was not told about, on a living
  subject carrying a cause of death, and on `cause_unknown_to = 0`, which would
  censor a dead subject in one outcome column while the other records a death.
- `fit_km()` stops on a requested time point past a stratum's follow-up, unless
  told to return `NA` there instead.
- `fit_cox()` stops when the proportional-hazards test cannot be computed, and
  drops a declared factor level that no row has rather than fitting an all-zero
  design column.
- `calc_rmst()` stops on a `tau` past the follow-up of either arm. The boundary
  is inclusive: `tau` exactly equal to the largest usable value is estimated.
- `fit_competing()` re-checks the two outcome columns against each other before
  anything is fitted, so a pair that has drifted apart -- each column still
  valid on its own -- is caught.

```{r refuse, error = TRUE}
prep_surv(sim_gist, time = "time_mo", status = "vital_status",
          cause = "cause_of_death",
          cause_gist_value  = "Dead (attributable to this cancer dx)",
          cause_other_value = "Other cause")
```

The last call fails because `"Dead (missing/unknown COD)"` was never declared.
That is the intended behaviour: with the set of admissible causes closed, a
code the analysis has not accounted for is an error, not a silent recode.

## Session

```{r session}
sessionInfo()
```
